Formalization of Typed Lua

### 2015-06-10 doc/papers/paper/paper.tex
rewrite of section on projection types

\subsection{Projections}
\label{sec:projections}

Lua programmers often overload the return type of functions to denote errors,
and our type system uses projection types to handle this idiom.

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
The function \emph{idiv} performs integer division and has type
\begin{align*}
& \Integer \times \Integer \times \Void \rightarrow \\
& (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
\end{align*}
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that describes
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Void$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(1, 2) \, \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \, q \, \mathbf{then} \, \lfloor print(q + r) \rfloor_{0} \, \mathbf{else} \, \lfloor print(``ERROR: " \, .. \, r) \rfloor_{0}$}
\end{tabular}
\end{center}

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$
for type checking $idiv(1, 2)$.
This relation means that given a type environment $\env_{1}$ and
a projection environment $\penv$, we can check that an expression
list $el$ has type $S_{1}$ and produces a new type environment $\env_{2}$
and produces a pair $(x,S_{2})$.
This pair means that the last expression of an expression list $el$
produces an union of second-level types $S_{2}$ that should be bound
to a variable $x$ in the projection environment $\penv$,
as the resulting type of this expression is a tuple of projection
types $\pi_{i}^{x}$.
In our example, our type system uses rule \textsc{T-EXPLIST4} for
type checking $idiv(1, 2)$:
\[
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:S, \env_{n+2}\\
       S = T_{n+1} \times ... \times T_{n+m} \times \Void \sqcup T_{n+1}' \times ... \times T_{n+m}' \times \Void\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times \Nil{*}, \env_{f}, (x,S)}
\end{array}
\]

Note that $idiv(1, 2)$ has type
$\pi_{1}^{x} \times \pi_{2}^{x} \times \Nil{*}$ and produces the pair
$(x,(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}))$.

In the rule that type checks the declaration of unannotated variables,
our type system uses the pair $(x,S_{2})$ to bound a union of
second-level types $S_{2}$ to a variable $x$ in the projection
environment $\penv$.
In our example, declaring $q$ and $r$
bounds the projection type $\pi_{1}^{x}$ to $q$ and
bounds the projection type $\pi_{2}^{x}$ to $r$,
where the projection variable $x$ bounds to 
\[
(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
in the projection environment $\penv$.
Rule \textsc{T-LOCAL2} illustrates this intuition:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{3} = \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)]\\
       \env_{3}, \penv[x \mapsto S_{2}] \vdash s, \env_{4} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{4} - \{\overline{id}\}}
\end{array}
\]

This rule uses the auxiliary function \emph{infer} to get the
most general first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.
We can define \emph{infer} as follows:
\begin{align*}
infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
\begin{array}{ll}
general(T_{i}) & \text{if $i < n$}\\
general(T_{n} \cup \Nil) & \text{if $i >= n$}
\end{array} \right.
\end{align*}

\begin{align*}
general(\False) & = \Boolean\\
general(\True) & = \Boolean\\
general({\it int}) & = \Integer\\
general({\it float}) & = \Number\\
general({\it string}) & = \String\\
general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
general(\{K_{1}{:}V_{1}, ..., \\
K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., \\
& \;\;\;\;\;\; K_{n}{:}general(V_{n})\}_{tag} \\
general(\mu x.T) & = \mu x.general(T)\\
general(T) & = T\\
\\
general2(\Void) & = \Void\\
general2(T{*}) & = general(T){*}\\
general2(T \times P) & = general(T) \times general2(P)\\
general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
\end{align*}

After assigning projection types to $q$ and $r$, reading $q$ will
use the projection type $\pi_{1}^{x}$ to project the type of $q$
into the union type $\Integer \cup \Nil$, while reading $r$ will
use the projection type $\pi_{2}^{x}$ to project the type of $r$
into the union type $\Integer \cup \String$.
Now, we may want to discriminate these variables to check whether
the function call returned with success.

Introducing a projection variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
\end{array}
\]

Rule \textsc{T-IF3} uses the auxiliary functions \emph{fpt} and \emph{gpt}
to filter a projection $x$, affecting all variables that bind to the same projection.
For instance, our previous example type checks through rule \textsc{T-IF3},
because it uses the information provided by the projection type $\pi_{1}^{x}$,
which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
\[
fpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
and the function call
\[
gpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.

>>>>

\subsection{Projection Types}
\label{sec:projections}

As we have seen in Section~\ref{sec:statistics}, Lua
programmers often overload the return type of functions
to denote errors, returning a {\tt nil} result and
an error message in case of an error instead of the
usual return values.

As an example, consider the integer division function
below, which returns the quotient and rest of integer
division, or a division by zero error:
\begin{verbatim}
  local function idiv(x : integer, y : integer)
    if divisor == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end
\end{verbatim}

Typed Lua infers the return type of this function as
$(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$. The idiom for using this such
a function is to test the first returned value:
\begin{verbatim}
  local q, r = idiv(a, b)
  if q then
    -- assume q and r are integers
  else 
    -- assume q is nil r is a string
  end
\end{verbatim}

A standard way of decomposing the union in the assignment
would let the type of {\tt q} be $\Integer \times \Nil$ and
the type of {\tt r} be $\Integer \times \String$. Flow typing
would narrow the type of {\tt q} to $\Integer$ in the {\tt then}
branch and to $\Nil$ in the {\tt else} branch, but would
leave the type of {\tt r} unchanged.

{\em Projection types}
are a general way to make the dependency between the types
of {\tt q} and {\tt r} survive the decomposition of the tuple,
so flow typing can narrow all of the components by testing
just one of them.

Intuitively, a projection type $\pi_{i}^{x}$ denotes the union of
the i-th components of all the tuples in the union of tuple types
denoted by the label $x$. Unions of tuple types referenced by projections are kept in their own {\em projection environment} $\Pi$. A fresh label is generated whenever the last term
in an expression list has a union of tuple types as its type.

Before introducing the typing rules for projections, let us translate the example above into the abstract syntax of Figure~\ref{fig:syntax}:
\begin{center}
	\begin{tabular}{ll}
		\multicolumn{2}{l}{$\mathbf{local} \, q, r = idiv(1, 2) \, \mathbf{in} \, \mathbf{if} \, q \, \mathbf{then} \, s_1 \, \mathbf{else} \, s_2$}
	\end{tabular}
\end{center}

Notice that the local variables $q$ and $r$ are declared without 
an explicit type. The rule for typing this {\bf local} statement,
{\sc T-LOCAL-UN}, introduces a new label in the projection
environment used for typing the {\bf if} statement:
\[
\begin{array}{c}
\mylabel{T-LOCAL-UN}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{4} - \{\overline{id}\}}
\end{array}
\]

This first sequent above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
This rule does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type,
as rule {\sc T-EXPLIST-PROJ} below shows:
\[
\begin{array}{c}
\mylabel{T-EXPLIST-PROJ}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
\end{array}
\]

In our example, the two rules combined will bind $x$ to
$(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})$ in the projection environment $\Pi$,
and $q$ and $r$ respectively to $\pi_{1}^{x}$ and $\pi_{2}^{x}$
in the type environment used for typing the {\bf if} statement.

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
clauses of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
\[
\begin{array}{c}
\mylabel{T-IF-FNIL-PROJ}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
\end{array}
\]

Rule \textsc{T-IF-FNIL-PROJ} uses the auxiliary functions \emph{fpt} and \emph{gpt}
to filter a projection $x$. Function {\em fpt} filters out $\Nil$ from the i-th
component of each tuple, removing the entire tuple if the i-th
component does not exist, or is $\Nil$ itself. Function {\em gpt}
filters the i-th component to $\Nil$ if the type includes
$\Nil$, removing the entire tuple if the i-th component does
not exist, or does not include $\Nil$.

In our example, $x$ would be bound to $\Integer \times \Integer \times \Nil{*}$ in the {\bf then} branch, and would be bound
to $\Nil \times \String \times \Nil{*}$ inside the {\bf else}
branch.

Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
